Classless Java

### 2016-03-24 OOPSLA16/sections/Conclusion.tex
13 pages

Before Java 8, concrete methods and static methods were not allowed
to appear in interfaces.  Java 8 allows static interface methods and
introduces \emph{default methods}, which enables implementations
inside interfaces. This had an important positive consequence that
was probably overlooked: the concept of class
(in Java) is now (almost) redundant and unneeded.
We proposed a programming style, called ClassLess Java, where
truly \objectoriented programs and (reusable) libraries
can be defined and used without ever defining a single class.

However, using this programming style directly in Java is very verbose.
To avoid syntactic boilerplate
caused by Java not being originally designed to work without classes,
we introduce a new annotation, \mixin, which provides default implementations
for various methods (e.g. getters, setters, with-methods) and a
mechanism to instantiate objects. 
We leverage on annotation processing and the Lombok library, in this way
\mixin is just a normal Java library; thus our proposed style can be integrated
in any Java project.

The \mixin annotation helps programmers
to write less cumbersome code while coding in Classless Java. Indeed, 
we think the obtained gain is so high that Classless Java with the \mixin
annotation can be less cumbersome than full Java.

Class less Java is just a programming style, but is 
showing the way for a new flavour of object orientation:
We propose \interfacebased \objectoriented languages (IB),
as opposed to \classbased or \prototypebased.

In IB state is not modelled at the platonic/ideal level
but is handled excursively by instances.

This unlock useful code reuse patterns, as shown in~\ref{sec:ep}.

Interestingly, without classes there is also no subclassing. This scratches an old
  itching point in the long struggle of subtyping versus subclassing:
  according to some authors, from a software engineering perspective,
  interfaces are just a kind of classes. Others consider more
  opportune to consider interfaces as pure types. We do not know how to conciliate
  those two viewpoints and Classless Java design.
  Classless Java does not have classes purely in the Java sense.  
% Classless Java encourages coding in a more flexible way by either
% keeping a higher abstraction level (interfaces are a more abstract
% concept than classes), or relying on concrete object initialization
% (the \Q@new I(){...}@ construct).

>>>>

Before Java 8, concrete methods and static methods were not allowed
to appear in interfaces.  Java 8 allows static interface methods and
introduces \emph{default methods}, which enables implementations
inside interfaces. This had an important positive consequence that
was probably overlooked: the concept of class
(in Java) is now (almost) redundant and unneeded.
We proposed a programming style, called Classless Java, where
truly \objectoriented programs and (reusable) libraries
can be defined and used without ever defining a single class.

However, using this programming style directly in Java is very verbose.
To avoid syntactic boilerplate
caused by Java not being originally designed to work without classes,
we introduce a new annotation, \mixin, which provides default implementations
for various methods (e.g. getters, setters, with-methods) and a
mechanism to instantiate objects. 
We leverage on annotation processing and the Lombok library, in this way
\mixin is just a normal Java library; thus our proposed style can be integrated
in any Java project.
The \mixin annotation helps programmers
to write less cumbersome code while coding in Classless Java. Indeed, 
we think the obtained gain is so high that Classless Java with the \mixin
annotation can be less cumbersome than full Java.

Classless Java is just a programming style, but is showing the way
for a new flavour of object orientation: We propose \interfacebased
\objectoriented languages (IB), as opposed to \classbased or
\prototypebased.  In IB state is not modelled at the platonic/ideal
level but is handled excursively by instances.  This unlocks useful
code reuse patterns, as shown in Section~\ref{sec:ep}.
An interesting avenue for future work would be to design a new
language based on IB. With a proper language design we would not need 
to restrict ourselfves to the limitations of Java and its syntax.

%Interestingly, without classes there is also no subclassing. This scratches an old
% itching point in the long struggle of subtyping versus subclassing:
%  according to some authors, from a software engineering perspective,
%  interfaces are just a kind of classes. Others consider more
%  opportune to consider interfaces as pure types. We do not know how to conciliate
%  those two viewpoints and Classless Java design.
%  Classless Java does not have classes purely in the Java sense.

% Classless Java encourages coding in a more flexible way by either
% keeping a higher abstraction level (interfaces are a more abstract
% concept than classes), or relying on concrete object initialization
% (the \Q@new I(){...}@ construct).

### 2016-03-24 OOPSLA16/sections/RelatedWork.tex
13 pages

%\paragraph{C++ Approach.}
C++ tries to provide a general solution to multiple inheritance by
virtual inheritance, dealing with the diamond problem by keeping only
one copy of the base class~\cite{ellis1990annotated}. However C++
suffers from the object initialization problem~\cite{malayeri2009cz}.
It bypasses all constructor calls to virtual superclasses, which can
potentially cause serious semantic errors. In our approach, the \mixin
annotation has full control over object initialization,
and the mechanism is transparent to users. Moreover, customized factory
methods are also allowed: if users are not satisfied with the default
generated \texttt{of} method, they can implement their own.

@@

%\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which
aims to do multiple inheritance without the diamond problem. They
divide inheritance into two separate concepts: inheritance dependency
and implementation inheritance. Using a combination of
\texttt{requires} and \texttt{extends}, a program with diamond
inheritance can be transformed to one without diamonds. Moreover,
fields and multiple inheritance can coexist. However untangling
inheritance also untangles the class structure. In CZ, not only the
number of classes, but also the class hierarchy complexity
increases. IB does not complicate the hierarchical structure, and state
can also coexist with multiple inheritance.

@@

\paragraph{Traits vs Object Interfaces.}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation-driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with state is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
state, while still being quite expressive.  Moreover the ability to support
constructors adds expressivity, which is not available in approaches
such as Scala's traits/mixins.\footnote{Scala traits have
  no constructor, but \Q@var@s and (\Q@lazy@) \Q@val@s can be
  ``pre-initialized''.}

@@

\paragraph{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namely it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolishes most advantages of interfaces.}
Object interfaces support automatic type-refinement.
Type refinement is part of a bigger topic in \classbased languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in \classbased \objectoriented
programming, and is exasperated by the \interfacebased programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
\objectoriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But this approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Our approach to covariantly refine state can recover
some of the additional expressiveness of MyType. As illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications, yet they do not require additional
complexity from the type system.

>>>>

%\paragraph{C++ Approach.}
C++ has a general solution to multiple inheritance by
virtual inheritance, dealing with the diamond problem by keeping only
one copy of the base class~\cite{ellis1990annotated}. However C++
suffers from the object initialization problem~\cite{malayeri2009cz}.
It bypasses constructor calls to virtual superclasses, which can
cause serious semantic errors. In our approach, the \mixin
annotation has full control over object initialization,
and the mechanism is transparent to users. Moreover, customized factory
methods are also allowed: if users are not satisfied with the default
generated \texttt{of} method, they can implement their own.

@@

%\paragraph{CZ.}
Malayeri and Aldrich proposed a model CZ~\cite{malayeri2009cz} which
aims to do multiple inheritance without the diamond problem. 
Inheritance is divided into two concepts: inheritance dependency
and implementation inheritance. Using a combination of
\texttt{requires} and \texttt{extends}, a program with diamond
inheritance is transformed to one without diamonds. Moreover,
fields and multiple inheritance can coexist. However untangling
inheritance also untangles the class structure. In CZ, not only the
number of classes, but also the class hierarchy complexity
increases. IB does not complicate the hierarchical structure, and state
also coexists with multiple inheritance.

@@

\paragraph{Traits vs Object Interfaces.}
We consider object interfaces to be an alternative to traits or mixins.  In the
trait model two concepts (traits and classes) coexist and cooperate. Some
authors~\cite{BettiniDSS13} see this as good language design fostering good
software development by helping programmers to think about the structure of
their programs.  However, other authors see the need of two concepts and the
absence of state as drawbacks of this model~\cite{malayeri2009cz}.  Object
interfaces are units of reuse, and at the same time they provide factory methods
for instantiation and support state.  Our approach promotes the use of
interfaces instead of classes, in order to rely on the modular composition
offered by interfaces. Since Java was designed for classes, a direct classless
programming style is verbose and unnatural. However, annotation-driven code
generation is enough to overcome this difficulty, and the resulting programming
style encourages modularity, composability and reusability, by keeping a strong
OO feel. In that sense, we promote object interfaces as being both units of
reusable code and object factories. Our practical experience is that, in Java,
separating the two notions leads to a lot of boilerplate code, and is quite
limiting when multiple inheritance with state is required.  Abstract state
operations avoid the key difficulties associated with multiple inheritance and
state, while still being quite expressive.  Moreover the ability to support
constructors adds expressivity, which is not available in approaches
such as Scala's traits/mixins.
%\footnote{Scala traits have
%  no constructor, but \Q@var@s and (\Q@lazy@) \Q@val@s can be
%  ``pre-initialized''.}

@@

\paragraph{ThisType and MyType}
%\marco{ this section is unfair. There are unsolved problems with ThisType in negative positions, namely it breaks subtyping. The most common solution is to allow calling some methods only when the exact type is known. This demolishes most advantages of interfaces.}
Object interfaces support automatic type-refinement.
Type refinement is part of a bigger topic in \classbased languages: expressing and
preserving type recursion and (nominal/structural) subtyping at the same time.
One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance. However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is a big limitation in \classbased \objectoriented
programming, and is exasperated by the \interfacebased programming we propose: no
type is ever going to be exact since classes are not explicitly used. A recent
article~\cite{Saito2013933} lights up this topic, proposing two
new features: exact statements and nonheritable methods. Both are
related to our work: 1) any method generated inside the \Q@of@ method is indeed
non-inheritable since there is no class name to extend from; 2) exact
statements (a form of wild-card capture on the exact run-time type) could
capture the ``exact type'' of an object even in a class-less
environment.
Admittedly, MyType greatly enhances the expressiveness and extensibility of
\objectoriented programming languages. Object interfaces use covariant-return types to
simulate some uses of MyType. But this approach only works for refining
return types, whereas MyType is more general, as it also works for
parameter types. Our approach to covariantly refine state can recover
some of the additional expressiveness of MyType. As illustrated with our examples and
case studies, object interfaces are still very useful in many
practical applications, yet they do not require additional
complexity from the type system.
